ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-26 pyked.tex
Update examples

The \ck{} file that will be used in this example can be found in the
\python|tests| directory of the PyKED repository,
\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself
%
\begin{minted}{yaml}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
fake-key: 0 # Again here because Atom's syntax highlighting is busted :(
\end{minted}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the data point is from the work of \textcite{Mittal2006a}:
%
\begin{minted}{yaml}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
  fake-key: 0 # Again here because Atom's syntax highlighting is busted :(
\end{minted}

Finally, there is just a single datapoint in this file, which codes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure,
%
\begin{minted}{yaml}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        - 0 # Dammit Atom
        ...
\end{minted}

The values for the \yaml|volume-history| are truncated to save space. The user
can load this file into \pk{} and run a simulation in Cantera with the following
code:

\begin{minted}{python}
import cantera as ct
from pyked import ChemKED

ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]

T_0 = dp.temperature.to('K').magnitude
P_0 = dp.pressure.to('Pa').magnitude
X_0 = dp.get_cantera_mole_fraction()

gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_0, P_0, X_0

reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))

time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

netw = ct.ReactorNet([reac])
while netw.time < 0.05:
    netw.step()

\end{minted}

@@

The \ck{} file used in this example can be found online.\todo{add url}{}
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol.
Examining the file, we first find the section that specifies the information about the \ck{} file itself:
%
\begin{minted}{yaml}
file-author:
  name: Morgan Mayer
  ORCID: 0000-0001-7137-5721
file-version: (0,1)
chemked-version: 0.0.1
screw-you-atom: 0
\end{minted}
%
Then the reference information:
%
\begin{minted}{yaml}
reference:
  doi: 10.1016/j.combustflame.2011.08.014
  authors:
    - name: Ivo Stranic
    - name: Deanna P. Chase
    - name: Joseph T. Harmon
    - name: Sheng Yang
    - name: David F. Davidson
    - name: Ronald K. Hanson
  journal: Combustion and Flame
  year: 2011
  volume: 159
  pages: 516-527
  detail:
experiment-type: ignition delay
apparatus:
  kind: shock tube
  institution: High Temperature Gasdynamics Laboratory, Stanford University
  facility: stainless steel shock tube
dammit-atom: 0
\end{minted}
%
In this \ck{} file, multiple data points are specified with some common
conditions, including a common mixture composition and common defintion of
ignition delay. Therefore, a \yaml|common-properties| section is specified.
%
\begin{minted}{yaml}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
atom-is-bad-at-syntax-highlighting: 0
\end{minted}
%
This block utilizes the ability of YAML files to define an anchor using the
\yaml|&| symbol and refer to that section later using the \yaml|*| symbol, as
can be seen in the definition of the \yaml|datapoints| section:
%
\begin{minted}{yaml}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{minted}

In this example, we would like to run constant volume simulations at each of the
pressure and temperature conditions in the \yaml|datapoints| list.

\begin{minted}{python}
import cantera as ct
from pyked import ChemKED

@@

X_0 = ck.datapoints[0].get_cantera_mole_fraction()

for dp in ck.datapoints:
    T_0 = dp.temperature.to('K').magnitude
    P_0 = dp.pressure.to('Pa').magnitude
    gas.TPX = T_0, P_0, X_0
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    while reac.T < T_0 + 400:
        netw.step()

    print('The ignition delay for T_0={}, P_0={} is: {} seconds'.format(
        T_0, P_0, netw.time
    ))
\end{minted}
%====================================================================
\section{Conclusions and future work}
%
\todo[inline]{both to do}

>>>>

The \ck{} file that will be used in this example can be found in the
\python|tests| directory of the PyKED repository,
\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the data point is from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which codes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yaml|volume-history| are truncated to save space. A user
might like to load the information from this file to perform a simulation to
calculate the ignition delay using Cantera. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic mechanism for H$_2$/CO combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \python|DataPoint| class. \pk{} uses Pint \python|Quantities| to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_0 = dp.temperature.to('K').magnitude
P_0 = dp.pressure.to('Pa').magnitude
X_0 = dp.get_cantera_mole_fraction()

# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_0, P_0, X_0

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \python|IdealGasReactor|, a
\python|Wall| is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\python|cansen.profile.VolumeProfile| class.
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

netw = ct.ReactorNet([reac])

# Integrate for 50 ms
while netw.time < 0.05:
    netw.step()
\end{pythonbox}
%
Although not shown in this example, the user would typically store information
about the \python|IdealGasReactor| (e.g., temperature, pressure, mass fractions)
at each time step of the integration for post-processing.

@@

The \ck{} file used in this example can be found online.\todo{add url}{} The
data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
In this \ck{} file, multiple data points are specified with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yaml|common-properties| section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block utilizes the ability of YAML files to define an anchor using the
\yaml|&| symbol and refer to that section later using the \yaml|*| symbol, as
can be seen in the definition of the \yaml|datapoints| section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}

In this example, we would like to run constant volume simulations at each of the
pressure and temperature conditions in the \yaml|datapoints| list. Once again,
the \ck{} file specifies all the information required for the simulations except
for the chemical kinetic mechanism.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

@@

# Since the composition is specified by the common-properties,
# just take the composition of the first DataPoint
X_0 = ck.datapoints[0].get_cantera_mole_fraction()

# Loop through each of the DataPoints
for dp in ck.datapoints:
    T_0 = dp.temperature.to('K').magnitude
    P_0 = dp.pressure.to('Pa').magnitude
    gas.TPX = T_0, P_0, X_0
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    # Define ignition delay as T_0 + 400 K
    while reac.T < T_0 + 400:
        netw.step()

    print('The ignition delay for T_0={}, P_0={} is: {} seconds'.format(
        T_0, P_0, netw.time
    ))
\end{pythonbox}
%
Note that the ignition delay in this example is defined as equal to $T_0 +
400\,\mathrm{K}$, a simplified definition used for this example. In general, the
user could post-process the concentration information from the simulation to
determine one-half the maximum OH* concentration to match the experimental
definition of ignition delay.
%====================================================================
\section{Conclusions and future work}
%
\todo[inline]{both to do}
